Program Transformations for miniKanren

### 2020-07-06 papers/aplas/text/evaluation.tex
remove subfig captions
compress even more
fix indent in listing 1.10
fix refs

\begin{figure}[!t]
  \begin{subfigure}[c]{0.35\textwidth}
    \centering
    \begin{tabular}{c||c||c}
      & forward & backward \\
      \hline\hline
      Original       & 34.8ms & 1.6ms \\ \hline
      \ecce          & 26.7ms & 1.7ms \\ \hline
      Ideal          & 26.9ms & 1.7ms \\ \hline
      ConsPD         & 47.6ms & 1.7ms
    \end{tabular}
    \caption{Evaluation results for doubleAppendo}
    \label{tbl:doubleApp}
  \end{subfigure}
  \hfill
  \begin{subfigure}[c]{0.6\textwidth}
    \includegraphics[width=\textwidth]{../data/doubleAppendo/da700log.pdf}
    \caption{???}
    %\label{tbl:doubleApp}
  \end{subfigure}
  \caption{???}
\end{figure}

The relation \lstinline{doubleAppend$^o$} concatenates three lists by a conjunction of two calls of the \lstinline{append$^o$} relation (see fig.~\ref{doubleApp}).
These two calls share a variable~--- an intermediate list \lstinline{ts}.
The list \lstinline{xs} is traversed twice to construct the result: first when \lstinline{ts} is constructed and then when \lstinline{ts} is traversed during the second call.
This double traversal negatively impacts the execution time, but it can be removed by deforestation.

The better implementation of this relation (see fig.~\ref{doubleApp:ideal}) does not traverse the first list twice.
It first recursively copies \lstinline{xs} into the beginning of \lstinline{res}, and only when \lstinline{xs} is fully consumed, it calls the \lstinline{append$^o$} to concatenate \lstinline{ys} and \lstinline{zs}.
This implementation is provided as the ideal implementation of \lstinline{doubleAppend$^o$} in~\cite{de1999conjunctive}.

@@

There is no significant difference in the execution time in the backward direction between different specialized versions (see table~\ref{tbl:doubleApp}).
However the execution time in the forward direction differs: the program generated by \ecce shows the same speedup as the ideal program, while conservative partial deduction slows the program down.
The reason behind the slowdown is the premature introduction of fresh variables which is significant for the input data of this size.

@@

The relation \lstinline{maxLenght$^o$} computes the maximal element of the input list and its length (see fig.~\ref{cpd:maxandlength}) by conjunction of two relation calls.
The list \lstinline{xs} is traversed twice: first to compute the maximum in the relation \lstinline{max$^o$} and then to compute the length in the relation \lstinline{length$^o$}.
The input list contain Peano numbers which are compared by relations less-or-equal (\lstinline{le$^o$}) and greater-than (\lstinline{gt$^o$}).
Note that the comparison relations have three arguments with the last one representing the result of comparison.

@@

\begin{figure}[!t]
  \begin{subfigure}[c]{0.4\textwidth}
    \centering
    \begin{tabular}{c||c}
      & [1..100] \\ \hline\hline
      Original         & 48.5ms  \\ \hline
      Ideal            & 32.2ms  \\ \hline
      Ideal removed    & 19.6ms  \\ \hline
      \ecce            & 23.7ms  \\ \hline
      ConsPD          & 35.7ms
    \end{tabular}

    \caption{Execution time of maxlengtho}
    \label{tbl:maxlen}
  \end{subfigure}
  \hfill
  \begin{subfigure}[c]{0.5\textwidth}
    \includegraphics[width=\textwidth]{../data/maxLengtho/max.pdf}
    \caption{???}
    %\label{tbl:doubleApp}
  \end{subfigure}
  \caption{???}
\end{figure}

The ideal implementation provided in the CPD literature~\cite{de1999conjunctive} is shown on fig.~\ref{ideal:maxandlength}.
This implementation traverses the input list \lstinline{xs} only once and can be generated by tupling.
Note, the disjunct which unifies the third argument with \lstinline{^false}\footnote{An arrow lifts ordinary values to the logic domain} in the implementation of both \lstinline{le$^o$} and \lstinline{gt$^o$} can never contribute into the result of \lstinline{maxLength$^o$} execution.
Such disjuncts are removed by both \ecce and conservative partial deduction, thus we consider two ideal implementions: one with the original comparison relations (\emph{Ideal}) and another in which these disjuncts are deleted (\emph{Ideal removed}).

We measured the execution time of \lstinline{maxLength$^o$ lst m l}, where \lstinline{lst} is a list of Peano numbers from $1$ to $100$.
The results are presented in table~\ref{tbl:maxlen}.

@@

One possible implementation of the evaluator is presented in fig.~\ref{eval:last}.
Here the relation \lstinline{elem$^o$ subst v res} unifies \lstinline{res} with the value of the variable \lstinline{v} in the list \lstinline{subst}.
The relations \lstinline{and$^o$}, \lstinline{or$^o$}, and \lstinline{not$^o$} encode corresponding boolean connectives.

@@

The conservative partial deduction first unfolds those calls which are selected according to the heuristics.
Since exploring the implementations of boolean connectives makes more sense, they are unfolded before recursive calls of \lstinline{eval$^o$}.
The way conservative partial deduction treats this program is the same as it treats the other implementation in which boolean connectives
are moved to the left, as shown in fig.~\ref{eval:fst}.
This program is easier for \ecce to specialize which demonstrates how unequal is the behaviour of CPD for similar programs.

@@

Depending on the way a relation is implemented, it may take a different number of driving steps to reach the point when any useful information is derived through its unfolding.
Partial deduction tries to unfold every relation call unless it is unsafe, but not all relation calls serve to restrict the search space and thus should be unfolded.
In the implementation of \lstinline{eval$^o$} boolean connectives can effectively restrict variables within the conjunctions and should be unfolded until they do.
But depending on the way they are implemented, the different number of driving steps should be performed for that.
The simplest way to implement these relations is by mimicking a truth tables as demonstrated by the implementation of \lstinline{not$^o$} in fig.~\ref{not:table}.
It is enough to unfold such relation calls once to derive useful information about variables.

@@

The other way to implement boolean connectives is to express them using a single basic boolean relation such as \lstinline{nand$^o$} which is, in turn, has a table-based
implementation (see fig.~\ref{not:nando}). It will take several sequential unfoldings to derive that variables \lstinline{v} and \lstinline{w} should
be \lstinline{^true} when considering a call \lstinline{and$^o$ v w ^true} implemented via a basic relation.
Conservative partial deduction drives the selected call until it derives useful substitutions for the variables involved while CPD with deterministic unfolding may fail to do so.

@@

These four implementations are very different from \ecce standpoint.
We measured the time necessary to generate $1000$ formulas over two variables which evaluate to \lstinline{^true} (averaged over 10 runs).
The results are presented in table~\ref{tbl:eval}.

\begin{figure}[!t]
  \centering
  \begin{subfigure}[c]{0.35\textwidth}
    \centering
    %% \begin{tabular}{c||c|c|c|c}
    %%   & FirstPlain & FirstNando & LastPlain & LastNando \\ \hline\hline
    %%   Original         & 14.8ms     & 14.5ms     & 7.4ms     & 10.1ms    \\ \hline
    %%   \ecce            & 16.0ms     & 22.3ms     & 14.0ms    & 14.9ms    \\ \hline
    %%   ConsPD          & 9.1ms      & 9.7ms      & 9.1ms     & 9.9ms     \\
    %% \end{tabular}
    \begin{tabular}{e{1cm}||c|c|c}
      & Original & \ecce & ConsPD \\ \hline\hline
      FirstPlain & 14.8ms & 16.0ms & 9.1ms \\ \hline
      FirstNando & 14.5ms & 22.3ms & 9.7ms \\ \hline
      LastPlain & 7.4ms & 14.0ms & 9.1ms \\ \hline
      LastNando & 10.1ms & 14.9ms & 9.9ms
    \end{tabular}
    \caption{Execution time of evalo}
    \label{tbl:eval}
  \end{subfigure}
  \hfill
  \begin{subfigure}[c]{0.6\textwidth}
    \includegraphics[width=\textwidth]{../data/propEval/prop.pdf}
    \caption{???}
  \end{subfigure}
  \caption{???}
\end{figure}

@@

\begin{table}
  \setlength{\tabcolsep}{0.5cm}
  \centering
  \begin{tabular}{c c}
    \infer[]{\Gamma \vdash IConst \ i : Int}{} &
    \infer[]{\Gamma \vdash BConst \ b : Bool}{} \vspace{0.5cm} \\

    \infer[]{\Gamma \vdash t + s : Int}{\Gamma \vdash t : Int, \Gamma \vdash  s : Int} &
    \infer[]{\Gamma \vdash t * s : Int}{\Gamma \vdash t : Int, \Gamma \vdash  s : Int} \vspace{0.5cm} \\

    \infer[]{\Gamma \vdash t = s : Bool}{\Gamma \vdash t : \tau, \Gamma \vdash  s : \tau} &
    \infer[]{\Gamma \vdash t < s : Bool}{\Gamma \vdash t : Int, \Gamma \vdash  s : Int} \vspace{0.5cm} \\

    \infer[]{\Gamma \vdash \underline{let} \ v \ b : \tau}{\Gamma \vdash v : \tau_v, \ (\tau_v :: \Gamma) \vdash b : \tau} &
    \infer[\Gamma \lbrack v \rbrack \equiv \tau]{\Gamma \vdash Var \ v : \tau}{} \vspace{0.5cm} \\

    \multicolumn{2}{c}{
      \infer[]{\Gamma \vdash \underline{if} \ c \ \underline{then} \ t \ \underline{else} \ s : \tau}{\Gamma \vdash c : Bool, \Gamma \vdash t : \tau, \Gamma \vdash s : \tau}
    }
  \end{tabular}
\end{table}

We compared two implementations of these typing rules.
The first one is obtained by unnesting of the functional program as described in~\cite{lozov2019relational} (\emph{Generated}).
The second program is hand-written in \oc (\emph{Hand-written}).
Each implementation has been specialized with ConsPD and \ecce.
We measured the time needed to generate 1000 closed terms of type integer (averaged over 10 iterations).

@@

Note that the generated program is far slower than hand-written.
The principal difference between these two implementations is that the generated program contains a certain redundancy.
For example, typechecking of the sum of two terms in the hand-written implementation is implemented by a single conjunction (see fig.~\ref{type:hand}) while the generated program is far more complicated and also uses a special relation \lstinline{typeEq$^o$} to compare types (see fig.~\ref{type:gen}).

@@

The relation \lstinline{isPath$^o$} was executed to search for 3 paths of length 7 in the graph with 20 vertices and 30 edges (see table~\ref{tbl:unify}).
Both specializers significantly improve the performance as compared to the original program, although ConsPD is about 30\% worse than \ecce.

>>>>

The relation \lstinline{doubleAppend$^o$} concatenates three lists by a conjunction of two calls of the \lstinline{append$^o$} relation (see Listing~\ref{doubleApp}).
These two calls share a variable~--- an intermediate list \lstinline{ts}.
The list \lstinline{xs} is traversed twice to construct the result: first when \lstinline{ts} is constructed and then when \lstinline{ts} is traversed during the second call.
This double traversal negatively impacts the execution time, but it can be removed by deforestation.

The better implementation of this relation (see Listing~\ref{doubleApp:ideal}) does not traverse the first list twice.
It first recursively copies \lstinline{xs} into the beginning of \lstinline{res}, and only when \lstinline{xs} is fully consumed, it calls the \lstinline{append$^o$} to concatenate \lstinline{ys} and \lstinline{zs}.
This implementation is provided as the ideal implementation of \lstinline{doubleAppend$^o$} in~\cite{de1999conjunctive}.

@@

There is no significant difference in the execution time in the backward direction between different specialized versions (see Fig.~\ref{tbl:doubleApp}).
However the execution time in the forward direction differs: the program generated by \ecce shows the same speedup as the ideal program, while conservative partial deduction slows the program down.
The reason behind the slowdown is the premature introduction of fresh variables which is significant for the input data of this size.

\begin{figure}[!t]
  \begin{subfigure}[c]{0.35\textwidth}
    \centering
    \begin{tabular}{c||c||c}
      & forward & backward \\
      \hline\hline
      Original       & 34.8ms & 1.6ms \\ \hline
      \ecce          & 26.7ms & 1.7ms \\ \hline
      Ideal          & 26.9ms & 1.7ms \\ \hline
      ConsPD         & 47.6ms & 1.7ms
    \end{tabular}
  \end{subfigure}
  \hfill
  \begin{subfigure}[c]{0.6\textwidth}
    \includegraphics[width=\textwidth]{../data/doubleAppendo/da700log.pdf}
  \end{subfigure}
    \caption{Evaluation results for doubleAppendo}
    \label{tbl:doubleApp}
\end{figure}

@@

@@

\begin{figure}[!h]
  \begin{subfigure}[c]{0.4\textwidth}
    \centering
    \begin{tabular}{c||c}
      & [1..100] \\ \hline\hline
      Original         & 48.5ms  \\ \hline
      Ideal            & 32.2ms  \\ \hline
      Ideal removed    & 19.6ms  \\ \hline
      \ecce            & 23.7ms  \\ \hline
      ConsPD          & 35.7ms
    \end{tabular}
  \end{subfigure}
  \hfill
  \begin{subfigure}[c]{0.45\textwidth}
    \includegraphics[width=\textwidth]{../data/maxLengtho/max.pdf}
  \end{subfigure}
  \caption{Execution time of maxlengtho}
  \label{tbl:maxlen}
\end{figure}

The relation \lstinline{maxLenght$^o$} computes the maximal element of the input list and its length (see Listing~\ref{cpd:maxandlength}) by conjunction of two relation calls.
The list \lstinline{xs} is traversed twice: first to compute the maximum in the relation \lstinline{max$^o$} and then to compute the length in the relation \lstinline{length$^o$}.
The input list contain Peano numbers which are compared by relations less-or-equal (\lstinline{le$^o$}) and greater-than (\lstinline{gt$^o$}).
Note that the comparison relations have three arguments with the last one representing the result of comparison.

The ideal implementation provided in the CPD literature~\cite{de1999conjunctive} is shown on Listing~\ref{ideal:maxandlength}.
This implementation traverses the input list \lstinline{xs} only once and can be generated by tupling.
Note, the disjunct which unifies the third argument with \lstinline{^false}\footnote{An arrow lifts ordinary values to the logic domain} in the implementation of both \lstinline{le$^o$} and \lstinline{gt$^o$} can never contribute into the result of \lstinline{maxLength$^o$} execution.
Such disjuncts are removed by both \ecce and conservative partial deduction, thus we consider two ideal implementions: one with the original comparison relations (\emph{Ideal}) and another in which these disjuncts are deleted (\emph{Ideal removed}).

We measured the execution time of \lstinline{maxLength$^o$ lst m l}, where \lstinline{lst} is a list of Peano numbers from $1$ to $100$.
The results are presented in Fig.~\ref{tbl:maxlen}.

@@

One possible implementation of the evaluator is presented in Listing~\ref{eval:last}.
Here the relation \lstinline{elem$^o$ subst v res} unifies \lstinline{res} with the value of the variable \lstinline{v} in the list \lstinline{subst}.
The relations \lstinline{and$^o$}, \lstinline{or$^o$}, and \lstinline{not$^o$} encode corresponding boolean connectives.

@@

The conservative partial deduction first unfolds those calls which are selected according to the heuristics.
Since exploring the implementations of boolean connectives makes more sense, they are unfolded before recursive calls of \lstinline{eval$^o$}.
The way conservative partial deduction treats this program is the same as it treats the other implementation in which boolean connectives
are moved to the left, as shown in Listing~\ref{eval:fst}.
This program is easier for \ecce to specialize which demonstrates how unequal is the behaviour of CPD for similar programs.

@@

Depending on the way a relation is implemented, it may take a different number of driving steps to reach the point when any useful information is derived through its unfolding.
Partial deduction tries to unfold every relation call unless it is unsafe, but not all relation calls serve to restrict the search space and thus should be unfolded.
In the implementation of \lstinline{eval$^o$} boolean connectives can effectively restrict variables within the conjunctions and should be unfolded until they do.
But depending on the way they are implemented, the different number of driving steps should be performed for that.
The simplest way to implement these relations is by mimicking a truth tables as demonstrated by the implementation of \lstinline{not$^o$} in Listing~\ref{not:table}.
It is enough to unfold such relation calls once to derive useful information about variables.

@@

The other way to implement boolean connectives is to express them using a single basic boolean relation such as \lstinline{nand$^o$} which is, in turn, has a table-based
implementation (see Listing~\ref{not:nando}). It will take several sequential unfoldings to derive that variables \lstinline{v} and \lstinline{w} should
be \lstinline{^true} when considering a call \lstinline{and$^o$ v w ^true} implemented via a basic relation.
Conservative partial deduction drives the selected call until it derives useful substitutions for the variables involved while CPD with deterministic unfolding may fail to do so.

@@

These four implementations are very different from \ecce standpoint.
We measured the time necessary to generate $1000$ formulas over two variables which evaluate to \lstinline{^true} (averaged over 10 runs).
The results are presented in Fig.~\ref{tbl:eval}.

\begin{figure}[!t]
  \centering
  \begin{subfigure}[c]{0.35\textwidth}
    \centering
    %% \begin{tabular}{c||c|c|c|c}
    %%   & FirstPlain & FirstNando & LastPlain & LastNando \\ \hline\hline
    %%   Original         & 14.8ms     & 14.5ms     & 7.4ms     & 10.1ms    \\ \hline
    %%   \ecce            & 16.0ms     & 22.3ms     & 14.0ms    & 14.9ms    \\ \hline
    %%   ConsPD          & 9.1ms      & 9.7ms      & 9.1ms     & 9.9ms     \\
    %% \end{tabular}
    \begin{tabular}{e{1cm}||c|c|c}
      & Original & \ecce & ConsPD \\ \hline\hline
      FirstPlain & 14.8ms & 16.0ms & 9.1ms \\ \hline
      FirstNando & 14.5ms & 22.3ms & 9.7ms \\ \hline
      LastPlain & 7.4ms & 14.0ms & 9.1ms \\ \hline
      LastNando & 10.1ms & 14.9ms & 9.9ms
    \end{tabular}
  \end{subfigure}
  \hfill
  \begin{subfigure}[c]{0.6\textwidth}
    \includegraphics[width=\textwidth]{../data/propEval/prop.pdf}
  \end{subfigure}
  \caption{Execution time of evalo}
  \label{tbl:eval}
\end{figure}

@@

\begin{figure*}[!h]
  \includegraphics[width=\textwidth]{typechecker-crop.pdf}
\end{figure*}

We compared two implementations of these typing rules.
The first one is obtained by unnesting of the functional program as described in~\cite{lozov2019relational} (\emph{Generated}).
The second program is hand-written in \oc (\emph{Hand-written}).
Each implementation has been specialized with ConsPD and \ecce.
We measured the time needed to generate 1000 closed terms of type integer (averaged over 10 iterations; see Table~\ref{tbl:eval}).

@@

Note that the generated program is far slower than hand-written.
The principal difference between these two implementations is that the generated program contains a certain redundancy.
For example, typechecking of the sum of two terms in the hand-written implementation is implemented by a single conjunction (see Listing~\ref{type:hand}) while the generated program is far more complicated and also uses a special relation \lstinline{typeEq$^o$} to compare types (see Listing~\ref{type:gen}).

@@

The relation \lstinline{isPath$^o$} was executed to search for 3 paths of length 7 in the graph with 20 vertices and 30 edges (see Table~\ref{tbl:unify}).
Both specializers significantly improve the performance as compared to the original program, although ConsPD is about 30\% worse than \ecce.
